Shallow EDSLs and Object-Oriented Programming

### 2016-06-24 paper/sections/Transformations.tex
Clean up code
Revise S4

\section{Transformations in Shallow Embeddings}
Transformations are typically regarded as the privilege of deep embeddings.
Since there is not a data structure persisting the AST, how to define transformations
becomes unclear.
It is not completely true, especially for the OO approach.
This section illustrates how to define various transformations, the missing
kind of interpretations in Gibbons and Wu's paper, in shallow embeddings.
%In fact, transformations can still be defined in shallow embedding.

\subsection{Desugaring}
Desugaring is one kind of transformations, which eliminates some
language constructs without sacrificing the expressiveness of the whole language.
In \dsl, the \emph{id} construct is such a syntactic sugar that can
be rewritten via the following formula:
$$id\ n = \overbrace{\ beside\ (fan\ 1)}^{\text{repeat }n-1\text{ times}}\ (fan\ 1)$$
which states that \emph{id n} can be represented as \emph{n} of \emph{fan 1} combined with \emph{beside}.
%Using the above formula, we can define a transformation that desugars every
%occurrence of \emph{identity} to a combination of \emph{fan} and \emph{beside}.

The following shows how to eliminate \lstinline{Id} construct in Scala:
\lstinputlisting[linerange=4-23]{../src/desugar/Circuit.scala}%APPLY:DESUGAR_SCALA
The newly introduced method \lstinline{desugar} returns a
\lstinline{Circuit} reflecting the nature of transformations. The implementation of \lstinline{desugar} for each case is
straightforward: \emph{fan n} returns itself; $beside\ c_1\ c_2$ returns a new
$beside$ with $c_1$ and $c_2$ desugared; \emph{id n} just mimics the formula.

%\weixin{define a show function in the and show the result after desugaring?}
Inspired by the Scala implementation, we can port it back to Haskell thanks to
the laziness of Haskell:
\lstinputlisting[linerange=6-21]{./code/desugar.hs}%APPLY:DESUGAR

\subsection{Sophisticated Optimizations}
Sophisticated optimizations may need to inspect multiple internal
representations of values.
Suppose we want to merge consecutive vertical wires:
$$
beside\ (id\ m)\ (id\ n) = id\ (m + n)
$$
This rewrite rule applies only when both $c_1$ and $c_2$ of $beside\ c_1\ c_2$
are $id$ and also accesses their inner value.
This kind of optimizations are hard to define in shallow embeddings since it
requires deep pattern matching on the AST.
Fortunately, Scala has built-in support for pattern matching on objects.
We can not simply decorate classes with \lstinline{case} modifier
as case class to case class extension is not supported.
Alternatively, we could manually define an \lstinline{unapply} method, a.k.a
\emph{extractor}, in the companion object:
%Here comes the implementation of the optimization:
\lstinputlisting[linerange=4-22]{../src/optimizations/Circuit.scala}%APPLY:MERGEIDS_SCALA

For other OO languages, pattern matching can still be simulated through \emph{test methods} or \emph{type test and type casts},
although they may not be as elegant as extractors~\cite{emir2007matching}.
Interested reader can refer to \ref{} for implementations using these approaches.

>>>>

\section{Transformations in Shallow Embeddings}
Transformations are typically regarded as the privilege of deep embeddings as no
persistent AST obscures how to define transformations in shallow embeddings.
The \textsc{Composite} pattern applied in our OO approach encodes the AST, making
it possible to implement transformations. More interestingly, we can port back
some simple transformations to Haskell.
This sections illustrates how to define transformations, the missing kind of interpretations in Gibbons and Wu's paper, in shallow embeddings.

\subsection{Desugaring}
Desugaring is an transformation, which eliminates some
language constructs without sacrificing the expressiveness of the language.
\emph{id} construct is such a syntactic sugar:
$$id\ n = \overbrace{\ beside\ (fan\ 1)}^{\text{repeat }n-1\text{ times}}\ (fan\ 1)$$
The formula states that \emph{id n} can be rewritten as \emph{n} of \emph{fan 1}
combined with \emph{beside}.

A corresponding Scala implementation is given below:
\lstinputlisting[linerange=4-17]{../src/desugar/Circuit.scala}%APPLY:DESUGAR_SCALA
The method \lstinline{desugar} returns a
\lstinline{Circuit} reflecting the nature of transformations.
Implementing it for each case is
straightforward: \emph{fan n} returns itself; $beside\ c_1\ c_2$ returns a new
$beside$ with $c_1$ and $c_2$ desugared; \emph{id n} just mimics the formula.
For constructing circuits easily, we define some companion objects with an \lstinline{apply} method implemented:
\lstinputlisting[linerange=20-26]{../src/desugar/Circuit.scala}%APPLY:COMPANION

%\weixin{define a show function in the and show the result after desugaring?}
We can port the implementation back to Haskell thanks to its laziness:
\lstinputlisting[linerange=6-21]{./code/desugar.hs}%APPLY:DESUGAR

\subsection{Sophisticated Optimizations}
Sophisticated optimizations may need to inspect multiple internal
representations of values, which imposes difficulty to shallow embeddings.
For instance, consecutive \emph{id}s put side by side can be merged together:
$$
beside\ (id\ m)\ (id\ n) = id\ (m + n)
$$
This rule applies to a $beside\ c_1\ c_2$ only when both $c_1$ and $c_2$ are
$id$, also their inner values are extracted for constructing the term
on the right hand side of the equation.
Fortunately, Scala has built-in support for pattern matching by decorating
classes with \lstinline{case} modifiers or defining extractor objects.
For extensibility reasons, case classes can not be used
because case class to case class extension is not supported.
Instead, we manually define \lstinline{unapply} methods in companion
objects:
\lstinputlisting[linerange=19-22]{../src/optimizations/Circuit.scala}%APPLY:EXTRACTOR
Then we can define this optimization:
\lstinputlisting[linerange=4-16]{../src/optimizations/Circuit.scala}%APPLY:MERGEIDS_SCALA

For other OO languages, pattern matching can still be simulated through \emph{test methods} or \emph{type test and type casts},
though they may not be as elegant as extractors~\cite{emir2007matching}.
%Interested reader can refer to \ref{} for implementations using these approaches.
